---
title: "NewWave example"
author: "Federico Agostinis"
date: "22/6/2020"
output: html_document
---

```{r}
library(splatter)
library(NewWave)
library(scran)
library(irlba)
library(Rtsne)
library(ggplot2)
library(mclust)
```

I am going to show how to use NewWave with example data. To do this I am going to generate some data using splatter.

```{r}
params <- newSplatParams()
N=1000
set.seed(1234)
data <- splatSimulateGroups(params,batchCells=c(N/2,N/2),
                           group.prob = rep(0.1,10),
                           de.prob = 0.2,
                           verbose = FALSE) 
```

Now we have a dataset with 1000 cells and 10000 genes, I will use only 1000.
NewWave takes as input raw data, not normalized.

```{r}
set.seed(12359)
data <- data[sample(1:nrow(data),1000)]
```

As you can see there is a variable called batch in the colData section.

```{r}
colData(data)
```

**IMPORTANT:** For batch effecr removal the batch variable must be a factor

```{r}
data$Batch <- as.factor(data$Batch)
```

We also have a variable called Group that represent the cell type labels.

We can see the how the cells are distributed between group and batch
```{r}
pca <- prcomp_irlba(t(counts(data)),n=10)
plot_data <-data.frame(Rtsne(pca$x)$Y)
```

```{r}
plot_data$batch <- data$Batch
plot_data$group <- data$Group
```

```{r}
ggplot(plot_data, aes(x=X1,y=X2,col=group, shape=batch))+ geom_point()
```

There is a clear batch effect between the cells.

Let's try to correct it.

# NewWave

The default options use 1 core and no mini-batch approaches, it don't use the verbose option but I set it TRUE for example purpose(not suggested with big dataset).
The default number of latent variables is 2, I will set it equal to 10.
```{r}
res <- newWave(data,X = "~Batch", K=10, verbose = T)
```
In order to make it faster you can increase the number of cores:
```{r}
res <- newWave(data,X = "~Batch", K=10, verbose = T, children=14)
```

Or use some mini-batch approaches:

```{r}
res <- newWave(data,X = "~Batch", verbose = T,K=10, children=14, n_gene_disp = 100, n_gene_par = 100, n_cell_par = 100)
```

Now I can use the latent dimension rapresentation for  visualization purpose:

```{r}
latent <- reducedDim(res)

tsne_latent <- data.frame(Rtsne(latent)$Y)
tsne_latent$batch <- data$Batch
tsne_latent$group <- data$Group
```

```{r}
ggplot(tsne_latent, aes(x=X1,y=X2,col=group, shape=batch))+ geom_point()
```

or for clustering:

```{r}
cluster <- kmeans(latent, 10)

adjustedRandIndex(cluster$cluster, data$Group)
```
