---
title: "vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo=FALSE}
library(splatter)
library(NewWave)
library(irlba)
library(Rtsne)
library(ggplot2)
library(mclust)
```

```{r}
library(splatter)
library(NewWave)
library(irlba)
library(Rtsne)
library(ggplot2)
library(mclust)
```
I am going to show how to use NewWave with example data.
To do this I am going to generate some data using splatter.

```{r}
params <- newSplatParams()
N=100
set.seed(1234)
data <- splatSimulateGroups(params,batchCells=c(N/2,N/2),
                           group.prob = rep(0.1,10),
                           de.prob = 0.2,
                           verbose = FALSE) 
```

Now we have a dataset with 1000 cells and 10000 genes, I will use only 1000.
NewWave takes as input raw data, not normalized.

```{r}
set.seed(12359)
data <- data[-which(rowSums(counts(data))==0),]
data <- data[sample(1:nrow(data),100),]
```

As you can see there is a variable called batch in the colData section.

```{r}
colData(data)
```

**IMPORTANT:** For batch effecr removal the batch variable must be a factor

```{r}
data$Batch <- as.factor(data$Batch)
```

We also have a variable called Group that represent the cell type labels.

We can see the how the cells are distributed between group and batch
```{r}
pca <- prcomp_irlba(t(counts(data)),n=10)
plot_data <-data.frame(Rtsne(pca$x)$Y)
```

```{r}
plot_data$batch <- data$Batch
plot_data$group <- data$Group
```

```{r}
ggplot(plot_data, aes(x=X1,y=X2,col=group, shape=batch))+ geom_point()
```

There is a clear batch effect between the cells.

Let's try to correct it.

# NewWave

The default options use 1 core and no mini-batch approaches, it don't use
the verbose option but I set it TRUE for example purpose(not suggested with
big dataset). The default number of latent variables is 2.
```{r}
res <- newWave(data,X = "~Batch", K=2, verbose = TRUE)
```
In order to make it faster you can increase the number of cores using "children" parameter:
```{r}
res2 <- newWave(data,X = "~Batch", K=2, verbose = TRUE, children=2)
```

Or use some mini-batch approaches:
    - n_gene_disp : Number of genes to use in the dispersion optimization
    - n_cell_par : Number of cells to use in the cells related parameters
        optimization
    - n_gene_par : Number of genes to use in the genes related parameters
        optimization

```{r}
res3 <- newWave(data,X = "~Batch", verbose = TRUE,K=2, children=2,
                n_gene_disp=100,  n_gene_par = 100, n_cell_par = 100,
                commondispersion = FALSE)
```

Now I can use the latent dimension rapresentation for visualization purpose:

```{r}
latent <- reducedDim(res)

tsne_latent <- data.frame(Rtsne(latent)$Y)
tsne_latent$batch <- data$Batch
tsne_latent$group <- data$Group
```

```{r}
ggplot(tsne_latent, aes(x=X1,y=X2,col=group, shape=batch))+ geom_point()
```

or for clustering:

```{r}
cluster <- kmeans(latent, 10)

adjustedRandIndex(cluster$cluster, data$Group)
```
